Palmer Archipelago Penguins Data in the palmerpenguins R Package - An Alternative to Anderson's \emph{Irises}

### 2021-01-27 penguins.Rmd
ah updates

 Edgar Anderson's famous 1935 *Iris* dataset, which exists in the base R **datasets** package as **iris**, is ubiquitous in statistics and data science education. Due to a lack of metadata, non-intuitive variables (e.g. Sepal.Width), its publication in (and inextricable connection with) Fisher's 1936 paper in *Annals of Eugenics*, and perfect structure (e.g. no missing values) unrepresentative of most real-world environmental data, a modern alternative to **iris** is needed. Penguin morphological measurements for three species of *Pygoscelis* penguins that breed on islands throughout the Palmer Archipelago, Antarctica, provide an approachable, charismatic and near drop-in replacement for the **iris** data. Here, we introduce the **palmerpenguins** R package and describe the included **penguins** data. We directly compare the two datasets for selected analyses to demonstrate that R users, in particular teachers and learners currently using the **iris** data, can switch to the Palmer Archipelago penguins for many use cases including data wrangling and visualization, correlation, regression, hypothesis testing, multivariate analysis (e.g. PCA), cluster analysis and classification (e.g. by k-means).

@@

@@

Both **iris** and **penguins** are in tidy format [ref] with each column denoting a single variable and each row containing measurements for a single *Iris* flower or penguin. The two datasets are comparable in size: dimensions (columns × rows) are 5 × 150 and 8 × 344 for **iris** and **penguins**, respectively, and sample sizes within species are similar (Table 1). Notably, sample sizes differ for the three penguin species, while sample sizes in **iris** are equal (n = 50 for each Iris species). Multiple factor variables in **penguins** (*species*, *island* and *sex*) along with *year* create additional opportunities for grouping, compared to the single factor (*species*) in **iris**.

>>>>

 Edgar Anderson's famous **iris** dataset, which contains size measurements for three species of *Iris* collected in 1935, exists in the base R **datasets** package as **iris** and is ubiquitous in statistics and data science materials. Due to a lack of metadata, non-intuitive variables (e.g. Sepal.Width), its publication in (and inextricable connection with) Fisher's 1936 paper in *Annals of Eugenics*, and perfect structure (e.g. no missing values) unrepresentative of most real-world environmental data, a modern alternative to **iris** is needed. Penguin morphological measurements for three species of *Pygoscelis* penguins that breed on islands throughout the Palmer Archipelago, Antarctica, provide an approachable, charismatic and near drop-in replacement for the **iris** data. Here, we introduce the **palmerpenguins** R package and describe the included **penguins** data. We directly compare the two datasets for selected analyses to demonstrate that R users, in particular teachers and learners currently using the **iris** data, can switch to the Palmer Archipelago penguins for many use cases including data wrangling and visualization, correlation, regression, hypothesis testing, multivariate analysis (e.g. PCA), cluster analysis and classification (e.g. by k-means).

@@

### Overview

```{r, echo = FALSE}
overview_compare <- tribble(
  ~Feature, ~iris, ~penguins,
  "Year(s) collected", "1935", "2007 - 2009",
  "Dimensions (col x row)", "5 x 150", "8 x 344",
  "Documentation", "minimal", "complete metadata",
  "Variable classes", "double (4), factor (1)", "double (2), int (3), factor (3)",
  "Missing values?", "no (n = 0; 0.0%)", "yes (n = 19; 0.7%)"
)

overview_compare %>% 
  kable(caption = "Overview comparison of penguins and iris dataset features and characteristics.")
```

@@

Both **iris** and **penguins** are in tidy format [@wickham_tidy_2014] with each column denoting a single variable and each row containing measurements for a single *Iris* flower or penguin. The two datasets are comparable in size: dimensions (columns × rows) are 5 × 150 and 8 × 344 for **iris** and **penguins**, respectively, and sample sizes within species are similar (Table 1). Notably, sample sizes differ for the three penguin species, while sample sizes in **iris** are equal (n = 50 for each Iris species). Multiple factor variables in **penguins** (*species*, *island* and *sex*) along with *year* create additional opportunities for grouping, compared to the single factor (*species*) in **iris**.

### 2021-01-27 penguins.tex
ah updates

\abstract{%
Edgar Anderson's famous 1935 \emph{Iris} dataset, which exists in the
base R \textbf{datasets} package as \textbf{iris}, is ubiquitous in
statistics and data science education. Due to a lack of metadata,
non-intuitive variables (e.g.~Sepal.Width), its publication in (and
inextricable connection with) Fisher's 1936 paper in \emph{Annals of
Eugenics}, and perfect structure (e.g.~no missing values)
unrepresentative of most real-world environmental data, a modern
alternative to \textbf{iris} is needed. Penguin morphological
measurements for three species of \emph{Pygoscelis} penguins that breed
on islands throughout the Palmer Archipelago, Antarctica, provide an
approachable, charismatic and near drop-in replacement for the
\textbf{iris} data. Here, we introduce the \textbf{palmerpenguins} R
package and describe the included \textbf{penguins} data. We directly
compare the two datasets for selected analyses to demonstrate that R
users, in particular teachers and learners currently using the
\textbf{iris} data, can switch to the Palmer Archipelago penguins for
many use cases including data wrangling and visualization, correlation,
regression, hypothesis testing, multivariate analysis (e.g.~PCA),
cluster analysis and classification (e.g.~by k-means).
}

@@

@@

Both \textbf{iris} and \textbf{penguins} are in tidy format {[}ref{]}
with each column denoting a single variable and each row containing
measurements for a single \emph{Iris} flower or penguin. The two
datasets are comparable in size: dimensions (columns × rows) are 5 × 150
and 8 × 344 for \textbf{iris} and \textbf{penguins}, respectively, and
sample sizes within species are similar (Table 1). Notably, sample sizes
differ for the three penguin species, while sample sizes in
\textbf{iris} are equal (n = 50 for each Iris species). Multiple factor
variables in \textbf{penguins} (\emph{species}, \emph{island} and
\emph{sex}) along with \emph{year} create additional opportunities for
grouping, compared to the single factor (\emph{species}) in
\textbf{iris}.

@@

\caption{\label{tab:unnamed-chunk-1}Grouped sample size for iris (by species; n = 150 total) and penguins (by species and sex; n = 344 total). Penguins can be further grouped by variables for island and study year.}
\centering
\begin{tabular}[t]{l|c|l|c|c|c}
\hline
\multicolumn{2}{c|}{Iris sample size (by species)} & \multicolumn{4}{c}{Penguin sample size (by species and sex)} \\
\cline{1-2} \cline{3-6}
Iris species & Sample size & Penguin species & Female & Male & NA\\
\hline
setosa & 50 & Adélie & 73 & 73 & 6\\
\hline
versicolor & 50 & chinstrap & 34 & 34 & 0\\
\hline
virginica & 50 & gentoo & 58 & 61 & 5\\
\hline
\end{tabular}
\end{table}

@@

\caption{\label{tab:unnamed-chunk-3}K-means cluster assignments by species based on penguin bill length (mm) and depth (mm), and iris petal length (cm) and width (cm).}
\centering
\begin{tabular}[t]{c|c|c|c|c|c|c|c}
\hline
\multicolumn{4}{c|}{Penguins cluster assignments} & \multicolumn{4}{c}{Iris cluster assignments} \\
\cline{1-4} \cline{5-8}
Cluster & Adélie & chinstrap & gentoo & Cluster & setosa & versicolor & virginica\\
\hline
1 & NA & 9 & 116 & 1 & NA & 48 & 4\\
\hline
2 & 4 & 54 & 6 & 2 & 50 & NA & NA\\
\hline
3 & 147 & 5 & 1 & 3 & NA & 2 & 46\\
\hline
\end{tabular}
\end{table}

>>>>

\abstract{%
Edgar Anderson's famous \textbf{iris} dataset, which contains size
measurements for three species of \emph{Iris} collected in 1935, exists
in the base R \textbf{datasets} package as \textbf{iris} and is
ubiquitous in statistics and data science materials. Due to a lack of
metadata, non-intuitive variables (e.g.~Sepal.Width), its publication in
(and inextricable connection with) Fisher's 1936 paper in \emph{Annals
of Eugenics}, and perfect structure (e.g.~no missing values)
unrepresentative of most real-world environmental data, a modern
alternative to \textbf{iris} is needed. Penguin morphological
measurements for three species of \emph{Pygoscelis} penguins that breed
on islands throughout the Palmer Archipelago, Antarctica, provide an
approachable, charismatic and near drop-in replacement for the
\textbf{iris} data. Here, we introduce the \textbf{palmerpenguins} R
package and describe the included \textbf{penguins} data. We directly
compare the two datasets for selected analyses to demonstrate that R
users, in particular teachers and learners currently using the
\textbf{iris} data, can switch to the Palmer Archipelago penguins for
many use cases including data wrangling and visualization, correlation,
regression, hypothesis testing, multivariate analysis (e.g.~PCA),
cluster analysis and classification (e.g.~by k-means).
}

@@

\hypertarget{overview}{%
\subsubsection{Overview}\label{overview}}

\caption{\label{tab:unnamed-chunk-1}Overview comparison of penguins and iris dataset features and characteristics.}
\centering
\begin{tabular}[t]{l|l|l}
\hline
Feature & iris & penguins\\
\hline
Year(s) collected & 1935 & 2007 - 2009\\
\hline
Dimensions (col x row) & 5 x 150 & 8 x 344\\
\hline
Documentation & minimal & complete metadata\\
\hline
Variable classes & double (4), factor (1) & double (2), int (3), factor (3)\\
\hline
Missing values? & no (n = 0; 0.0\%) & yes (n = 19; 0.7\%)\\
\hline
\end{tabular}
\end{table}

@@

Both \textbf{iris} and \textbf{penguins} are in tidy format
\citep{wickham_tidy_2014} with each column denoting a single variable
and each row containing measurements for a single \emph{Iris} flower or
penguin. The two datasets are comparable in size: dimensions (columns ×
rows) are 5 × 150 and 8 × 344 for \textbf{iris} and \textbf{penguins},
respectively, and sample sizes within species are similar (Table 1).
Notably, sample sizes differ for the three penguin species, while sample
sizes in \textbf{iris} are equal (n = 50 for each Iris species).
Multiple factor variables in \textbf{penguins} (\emph{species},
\emph{island} and \emph{sex}) along with \emph{year} create additional
opportunities for grouping, compared to the single factor
(\emph{species}) in \textbf{iris}.

@@

\caption{\label{tab:unnamed-chunk-2}Grouped sample size for iris (by species; n = 150 total) and penguins (by species and sex; n = 344 total). Penguins can be further grouped by variables for island and study year.}
\centering
\begin{tabular}[t]{l|c|l|c|c|c}
\hline
\multicolumn{2}{c|}{Iris sample size (by species)} & \multicolumn{4}{c}{Penguin sample size (by species and sex)} \\
\cline{1-2} \cline{3-6}
Iris species & Sample size & Penguin species & Female & Male & NA\\
\hline
setosa & 50 & Adélie & 73 & 73 & 6\\
\hline
versicolor & 50 & chinstrap & 34 & 34 & 0\\
\hline
virginica & 50 & gentoo & 58 & 61 & 5\\
\hline
\end{tabular}
\end{table}

@@

\caption{\label{tab:unnamed-chunk-4}K-means cluster assignments by species based on penguin bill length (mm) and depth (mm), and iris petal length (cm) and width (cm).}
\centering
\begin{tabular}[t]{c|c|c|c|c|c|c|c}
\hline
\multicolumn{4}{c|}{Penguins cluster assignments} & \multicolumn{4}{c}{Iris cluster assignments} \\
\cline{1-4} \cline{5-8}
Cluster & Adélie & chinstrap & gentoo & Cluster & setosa & versicolor & virginica\\
\hline
1 & 147 & 5 & 1 & 1 & NA & 2 & 46\\
\hline
2 & NA & 9 & 116 & 2 & NA & 48 & 4\\
\hline
3 & 4 & 54 & 6 & 3 & 50 & NA & NA\\
\hline
\end{tabular}
\end{table}
